---
title: "Mock Project Using the Maximov (2021) Amphipod Dataset"
author: "Muhammad Uzair Davids"
date: "2025-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Loading packages for the analysis
First, I load the packages I want to use to clean the data set. 

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```


## Introduction to dataset

The dataset used for this exercise is taken from the paper by [Alexy Maximov 
(2021)](https://doi.org/10.5061/dryad.0zpc866z6) titled: **Population dynamics 
of the glacial relict amphipod *Monoporeia affinis* in a subarctic lake**.   

```{r load dataset}
env <- read.csv("data/data_environmental.csv") # environmental variables
macro <- read.csv("data/data_macrozoobenthos.csv") # species abundance data 
pop <- read.csv("data/data_population.csv") # age classes of population
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
