---
title: "**Mock Project Using the Maximov (2021) Amphipod Dataset**"
author: "Muhammad Uzair Davids"
date: "2025-02-20"
output: html_document
---

## **Setting up R**

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

## **Loading packages for the analysis**
First, I load the packages I want to use to clean the data set. 

```{r load library}
library(tidyverse) # The main package we will be using to tidy our data
library(tidyr) # For the pivot longer function
library(dplyr) # To rename the columns
library(ggplot2) # To make the pretty graphs at the end
library(lubridate) # Will allow us to extract the year from date data later on
```

--- 

## **Loading the datasets**

The datasets used for this exercise is based on a dataset by [Alexy Maximov 
(2021)](https://doi.org/10.5061/dryad.0zpc866z6) titled: **Population dynamics 
of the glacial relict amphipod *Monoporeia affinis* in a subarctic lake**. This
data set was used for the paper titled: Population dynamics of the glacial 
relict amphipods in a subarctic lake: role of density-dependent and 
density-independent factors. 

There are three datasets: `data_environmental.csv` contains the data pertaining 
to the environmental variables, `data_macrzoobenthos.csv` contains data 
pertaining to the abundances of certain groups macrozoobenthos organisms, and
`data_population.csv` contains the data pertaining to the age structure of the 
*Monoporeia* spp. population. 

```{r load dataset}
# Load in environmental data
raw_env <- read.csv(
  "data/data_environmental.csv"
)

# Load in macrzoobenthos abundance data
raw_macro <- read.csv(
  "data/data_macrozoobenthos.csv"
) 

# Load in population abundance data
raw_pop <- read.csv(
  "data/data_population.csv"
)

# I have loaded all the data sets into their own variables.
```

## **Inspect the data**

Now that we've loaded the data sets, lets have a closer look at the variables 
and some of the values they contain.

For a summary of the data sets we can use the `summary()` function. 
 
```{r summarise environmental data}
# summary of the environmental data set
summary(raw_env)
```

```{r summarise macrozoobenthos data}
# summary of the macrozoobenthos data set
summary(raw_macro)
```

```{r summarise population data}
# summary of the population data set
summary(raw_pop)
```

We can also use the `head()` and `tail()` functions. `head()` will allow us to 
view the first few rows of the data, whereas `tail()` will allow us to view the 
last few rows.

```{r first and last rows of the environmental data}
# first few rows of the environmental data
head(raw_env)

# last few rows of the environmental data
tail(raw_env)
```

```{r first and last rows of macrozoobenthos data}
# first few rows of the macrozoobenthos data
head(raw_macro)

# last few rows of the macrozoobenthos data
tail(raw_macro)
```

```{r first and last rows of the population data}
# first few rows of the population data
head(raw_pop)

# last few rows of the population data
tail(raw_pop)
```

---

# **Tidying the data**

The first step of this exercise will be to ensure that the data we're using is
in *tidy format*. Ensuring that our data is in *tidy format* helps to make the 
workflow easier to follow, but also allows for more effective communication and 
reproductibility of research. 

At its simplest *tidy data* is data where:

1. Columns represent variables.

2. Rows are observations.

3. Cells contain only a single value.

4. The data is in *wide format*.

5. There are no gaps in the data. 

It should also be mentioned that these guidelines may not be the only things
which qualify as *tidy data*. The data sets I am using abides by most of these 
guidelines, except for two - it is **not** in *wide format*, and there are data 
missing. 

There are also other steps we can take to tidy the data further, such as 
renaming columns, which may help with the reproducible aspect of this 
assignment.

## **Removing missing and invalid data**

Gaps in data sets are considered to be in violation of *tidy format*. The 
environmental data set contains cells with "No data" values - which we saw 
after using the `head()` function above. To fix this, we will replace the 
"No data" message with `NA`. 

```{r removing missing data}
raw_env[raw_env == "No data"] <- NA # Replace "No data" with NA

# NB: I have overwritten the old raw_env file here. 
```

## **Removing redundant data**

Currently the macrozoobenthos and population datasets contain data which could
be considered as redundant - the "Mean annual abundance of all macrozoobenthos 
(individuals.m-2)" column is just the sum of mean annual abundance of all the
organisms, and the "Total population abundance (individuals.m-2)" column is the
sum of all age groups in the population - so we will remove them to make 
plotting the data easier. 

```{r remove redundant data}
# remove redundant columns from the macrozoobenthos dataset
raw_macro <- raw_macro %>% select(-`Macrozoobenthos..ind.m.2`)

# remove redundant columns from the population dataset
raw_pop <- raw_pop %>% select(-`N.population..ind.m.2`)

#NB: for both of these functions I've overwritten the raw data since we won't 
# be needing the old data. 
```

## **Rename columns**

Some of the column names in the data set are not easily understood, so I 
will change them to make them easier to understand.

```{r rename columns}
# Rename the environmental data columns
tidy_env <- raw_env %>% rename(
  "Mean Concentration of Chlorophyll-a (µg.L-1)" 
  = Clorophyll.a..microgram.L.1,
  "Standard Error (µg.L-1)" 
  = SE,
  "Mean Near-Bottom Temperature (°C)" 
  = T.mean,
  "Minimum Near-Bottom Temperature (°C)" 
  = T.min,
  "Maximum Near-Bottom Temperature (°C)" 
  = T.max
)

# Rename the macrozoobenthos abundance data columns
tidy_macro <- raw_macro %>% rename(
  "Mean annual abundance of Monoporeia affinis (individuals.m-2)" 
  = Monoporeia..ind.m.2,
  "Mean annual abundance of chironomids (individuals.m-2)" 
  = Chironomidae..ind.m.2,
  "Mean annual abundance of other taxa (individuals.m-2)" 
  = Others..ind.m.2
)

# Rename the population abundance data columns
tidy_pop <- raw_pop %>% rename(
  "Abundance of age group 0+ (individuals.m-2)"
  = N.0...ind.m.2,
  "Abundance of age group 1+ (individuals.m-2)"
  = N.1...ind.m.2,
  "Abundance of age group 2+ (individuals.m-2)"
  = N.2...ind.m.2,
  "Standard error (individuals.m-2)" 
  = SE
)

# Here I have created new variables for the renamed data.  
```

## **Pivot longer**

Currently, our data is in *wide format*, which violates the tidy format 
guidelines. So let's fix that using the `pivot_longer` function from the `tidyr` 
package. This will allow us to convert the data into *long format*.

```{r pivot longer}
# convert the environmental data to long format
long_env <- tidy_env %>%
  pivot_longer(
    cols = c(
      "Mean Near-Bottom Temperature (°C)",
      "Minimum Near-Bottom Temperature (°C)",
      "Maximum Near-Bottom Temperature (°C)"
    ),
    # Select the columns to pivot
    names_to = "Variable",
    # New column name for previous column headers
    values_to = "Value"
  )      # New column for data values

# convert the macrozoobenthos data to long format
long_macro <- tidy_macro %>%
  pivot_longer(
    cols = c(
      "Mean annual abundance of Monoporeia affinis (individuals.m-2)",
      "Mean annual abundance of chironomids (individuals.m-2)",
      "Mean annual abundance of other taxa (individuals.m-2)"
    ),
    # Select the columns to pivot
    names_to = "Variable",
    # New column name for previous column headers
    values_to = "Value"
  )      # New column for data values

# convert the population data to long format
long_pop <- tidy_pop %>%
  pivot_longer(
    cols = c(
      "Abundance of age group 0+ (individuals.m-2)",
      "Abundance of age group 1+ (individuals.m-2)",
      "Abundance of age group 2+ (individuals.m-2)"
    ),
    # Select the columns to pivot
    names_to = "Variable",
    # New column name for previous column headers
    values_to = "Value"
  )      # New column for data values

# I have created new variables for the long format versions of the data. 
```

---
## **Working with the tidied data**
Now that we've converted the data to *long format*, we can consider our data to
be in tidy format - so let's create a plot. For this part of the exercise I 
will just work with the tidied population data set.

Firtsly, the data set contains multiple dates, but I want to look at how the 
age structure of the *Monoporeia affinis* population changes over the years on
average. So let's extract the year from the date column and summarise the 
average number of individuals per age group per year. 

```{r extract the year and calculate the average}
# Convert Date column to Date format (if not already)
long_pop <- long_pop %>%
  mutate(Year = year(Date))  # Extract year from the Date column

# Summarizing data: Average number of individuals per age group per year
pop_summary <- long_pop %>%
  group_by(Year, Variable) %>%  # Group by Year and Age Group
  summarise(Avg_Individuals = mean(Value), .groups = 'drop')  # Average per year
```


## **Plot the tidied data**
Now that we've extracted the year form the `Date` column and calculated the 
average number of individuals per m² per age group per year,  let's plot it!

Since we're working with data which represents groups within a larger 
population, and looking at how they change over time, a stacked bar graph may 
work best.  

```{r stacked bar graph}
# Create the stacked bar plot
ggplot(pop_summary,
       aes(x = as.factor(Year), y = Avg_Individuals, fill = Variable)) +
  geom_col(position = "stack") +  # Stacked bars
  scale_fill_grey(start = 0.2, end = 0.8) +  # Monochrome gradient
  theme_minimal() +
  labs(
    title = "Figure 1: The population age structure of Monoporeia affinis
       between the year 2002 and 2019",
    x = "Year",
    y = "Average abundance of age groups (individuals per m²)",
    fill = "Age Group"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate date labels
```

---
