---
title: "Mock Project Using the Maximov (2021) Amphipod Dataset"
author: "Muhammad Uzair Davids"
date: "2025-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages for the analysis
First, I load the packages I want to use to clean the data set. 

```{r load library, include=FALSE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
```

## Introduction to dataset

The dataset used for this exercise is taken from the paper by [Alexy Maximov 
(2021)](https://doi.org/10.5061/dryad.0zpc866z6) titled: **Population dynamics 
of the glacial relict amphipod *Monoporeia affinis* in a subarctic lake**.   

```{r load dataset}
env <- read.csv("data/data_environmental.csv") # environmental variables
macro <- read.csv("data/data_macrozoobenthos.csv") # species abundance data
pop <- read.csv("data/data_population.csv") # age classes of population
```

## Rename columns

Some of the column names in the data set are not easily understood, so I 
will change them to make them easier to understand and read from the data set.

```{r rename columns}
# Rename the environmental data columns
env <- env %>% rename(
  "Mean Concentration of Chlorophyll-a (µg.L-1)" = Clorophyll.a..microgram.L.1,
  "Standard Error (µg.L-1)" = SE,
  "Mean Near-Bottom Temperature (°C)" = T.mean,
  "Minimum Near-Bottom Temperature (°C)" = T.min,
  "Maximum Near-Bottom Temperature (°C)" = T.max
)  

macro <- macro %>% rename(
  "Mean Concentration of Chlorophyll-a (µg.L-1)" = Clorophyll.a..microgram.L.1,
  "Standard Error (µg.L-1)" = SE,
  "Mean Near-Bottom Temperature (°C)" = T.mean,
  "Minimum Near-Bottom Temperature (°C)" = T.min,
  "Maximum Near-Bottom Temperature (°C)" = T.max
)  

env <- env %>% rename(
  "Mean Concentration of Chlorophyll-a (µg.L-1)" = Clorophyll.a..microgram.L.1,
  "Standard Error (µg.L-1)" = SE,
  "Mean Near-Bottom Temperature (°C)" = T.mean,
  "Minimum Near-Bottom Temperature (°C)" = T.min,
  "Maximum Near-Bottom Temperature (°C)" = T.max
)  

```
